Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-06-22 manuscript.Rmd
Fixed error with IM figure

processdata <- 1  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 48GB of storage)
cleanupraw <- 1  # if downloading raw data, delete once processed to save space

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [e.g. @Anstis1998; @Levelt1965; @Legge1981; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

Pupil size affects the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [e.g. @Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [e.g. @Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the pupil and perceptual pathways are effectively distinct, but hope to investigate this more directly in future neuroimaging work.

@@

In the steady-state literature, one hallmark of a nonlinear system that pools inputs is the presence of intermodulation responses at the sums and differences of the fundamental flicker frequencies [@Tsai2012]. In Figure \@ref(fig:IMfig) we plot the amplitude spectra of conditions from Experiment 1 in which the two eyes were stimulated at different frequencies (2Hz and 1.6Hz) but at the same contrast (48\%; these correspond to the binocular cross and dichoptic cross conditions in Figures \@ref(fig:pupildata)d,e and \@ref(fig:EEGdata)d,e). Consistent with the temporal properties of pupil responses and EEG activity, Figure \@ref(fig:IMfig)a reveals a strong intermodulation difference response at 0.4Hz (red dashed line), and Figure \@ref(fig:IMfig)b reveals an intermodulation sum response at 3.6Hz (red dashed line). The presence of these intermodulation terms is predicted by nonlinear gain control models of the type considered here [@Tsai2012; @Baker2017], and indicates that the processing of monocular flicker signals is not fully linear prior to the point at which they are combined across the eyes. Indeed, our model architecture makes specific predictions about the location of interocular suppression - it impacts before binocular combination, consistent with results from primate physiology [@].

@@

@@

# (download and) analyse individual participant data for replication expt
processdata <- 3

@@

  aspratio <- 10/15  # this is the aspect ratio of the output pdf
  imwidth <- 0.2
  xstart <- 0.85
  ystart <- 0.72
  rasterImage(stim,xstart,ystart,xstart+imwidth*aspratio,ystart+imwidth)

>>>>

processdata <- 1  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 48GB of storage)
cleanupraw <- 0  # if downloading raw data, delete once processed to save space

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry; see Figure \@ref(fig:BSratios)). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [e.g. @Anstis1998; @Levelt1965; @Legge1981; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

Pupil size affects the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [e.g. @Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [e.g. @Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the pupil and perceptual pathways are effectively distinct, but hope to investigate this more directly in future neuroimaging work. Using fMRI to simultaneously image cortical and subcortical brain regions will also allow us to check that the differences we report here are not a consequence of the different measurement techniques we used (pupillometry and EEG).

@@

In the steady-state literature, one hallmark of a nonlinear system that pools inputs is the presence of intermodulation responses at the sums and differences of the fundamental flicker frequencies [@Tsai2012]. In Figure \@ref(fig:IMfig) we plot the amplitude spectra of conditions from Experiment 1 in which the two eyes were stimulated at different frequencies (2Hz and 1.6Hz) but at the same contrast (48\%; these correspond to the binocular cross and dichoptic cross conditions in Figures \@ref(fig:pupildata)d,e and \@ref(fig:EEGdata)d,e). Consistent with the temporal properties of pupil responses and EEG activity, Figure \@ref(fig:IMfig)a reveals a strong intermodulation difference response at 0.4Hz (red dashed line), and Figure \@ref(fig:IMfig)b reveals an intermodulation sum response at 3.6Hz (red dashed line). The presence of these intermodulation terms is predicted by nonlinear gain control models of the type considered here [@Tsai2012; @Baker2017], and indicates that the processing of monocular flicker signals is not fully linear prior to the point at which they are combined across the eyes. Indeed, our model architecture makes specific predictions about the location of interocular suppression - it impacts before binocular combination, consistent with results from primate physiology [@Dougherty2019].

@@

dev.off()
}

@@

# (download and) analyse individual participant data for replication expt

@@

### 2023-06-22 manuscript.tex
Fixed error with IM figure

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. \protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Legge1981}{Legge and Rubin, 1981}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}; \protect\hyperlink{ref-Quaia2018}{Quaia et al., 2018}). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

Pupil size affects the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2\% of overall diameter (e.g. \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}). It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms (e.g. \protect\hyperlink{ref-Bradley2008}{Bradley et al., 2008}). We think it unlikely that these temporally coarse processes would have a differential effect on e.g.~monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the pupil and perceptual pathways are effectively distinct, but hope to investigate this more directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase `blobs' (\protect\hyperlink{ref-Horton1981}{Horton and Hubel, 1981}; \protect\hyperlink{ref-Livingstone1984}{Livingstone and Hubel, 1984}) are biased towards low spatial frequencies (\protect\hyperlink{ref-Edwards1995}{Edwards et al., 1995}; \protect\hyperlink{ref-Tootell1988}{Tootell et al., 1988}), and relatively insensitive to stimulus orientation (\protect\hyperlink{ref-Horton1981}{Horton and Hubel, 1981}; \protect\hyperlink{ref-Livingstone1984}{Livingstone and Hubel, 1984}; though see \protect\hyperlink{ref-Economides2011}{Economides et al., 2011}). As the blob regions are embedded within ocular dominance columns (\protect\hyperlink{ref-Horton1981}{Horton and Hubel, 1981}), they are also largely monocular (\protect\hyperlink{ref-Livingstone1984}{Livingstone and Hubel, 1984}; \protect\hyperlink{ref-Tychsen2004}{Tychsen et al., 2004}). More recent work has reported psychophysical evidence for unoriented chromatic (\protect\hyperlink{ref-Gheiratmand2013}{Gheiratmand et al., 2013}) and achromatic (\protect\hyperlink{ref-Meese2011}{Meese and Baker, 2011}) mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g.~Figures \ref{fig:BSratios} \& \ref{fig:TFdata}) owing to additivity of the electrophysiological response across independent monocular cells, rather than within binocular neurons. In the steady-state literature, one hallmark of a nonlinear system is the presence of intermodulation responses at the sums and differences of fundamental flicker frequencies (\protect\hyperlink{ref-Tsai2012}{Tsai et al., 2012}). In Figure \ref{fig:IMfig} we plot the amplitude spectra of conditions from Experiment 1 in which the two eyes were stimulated at different frequencies (2Hz and 1.6Hz) but at the same contrast (48\%; these correspond to the binocular cross and dichoptic cross conditions in Figures \ref{fig:pupildata}d,e and \ref{fig:EEGdata}d,e). Consistent with the temporal properties of pupil responses and EEG, Figure \ref{fig:IMfig}a reveals a strong intermodulation difference response at 0.4Hz (red dashed line), and Figure \ref{fig:IMfig}b reveals an intermodulation sum response at 3.6Hz (red dashed line). The presence of these intermodulation terms is predicted by nonlinear gain control models of the type considered here (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Tsai2012}{Tsai et al., 2012}), and indicates that the processing of monocular flicker signals is not fully linear prior to the point at which they are combined across the eyes.

@@

\caption{Summary of intermodulation responses in pupillometry (a) and EEG (b). The data are from the binocular cross and dichoptic cross conditions of Experiment 1, with a target contrast of 48\%. Vertical dashed lines indicate the fundamental flicker frequencies of 2Hz (F1; black) and 1.6Hz (F2; green), and the intermodulation difference (F1-F2 = 0.4Hz) and sum (F1+F2 = 3.6Hz) frequencies. Data are averaged across N = 30 participants, and shaded regions indicate ±1 standard error.}\label{fig:IMfig}
\end{figure}

@@

>>>>

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry; see Figure \ref{fig:BSratios}). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. \protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Legge1981}{Legge and Rubin, 1981}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}; \protect\hyperlink{ref-Quaia2018}{Quaia et al., 2018}). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

Pupil size affects the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2\% of overall diameter (e.g. \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}). It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms (e.g. \protect\hyperlink{ref-Bradley2008}{Bradley et al., 2008}). We think it unlikely that these temporally coarse processes would have a differential effect on e.g.~monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the pupil and perceptual pathways are effectively distinct, but hope to investigate this more directly in future neuroimaging work. Using fMRI to simultaneously image cortical and subcortical brain regions will also allow us to check that the differences we report here are not a consequence of the different measurement techniques we used (pupillometry and EEG).

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase `blobs' (\protect\hyperlink{ref-Horton1981}{Horton and Hubel, 1981}; \protect\hyperlink{ref-Livingstone1984}{Livingstone and Hubel, 1984}) are biased towards low spatial frequencies (\protect\hyperlink{ref-Edwards1995}{Edwards et al., 1995}; \protect\hyperlink{ref-Tootell1988}{Tootell et al., 1988}), and relatively insensitive to stimulus orientation (\protect\hyperlink{ref-Horton1981}{Horton and Hubel, 1981}; \protect\hyperlink{ref-Livingstone1984}{Livingstone and Hubel, 1984}; though see \protect\hyperlink{ref-Economides2011}{Economides et al., 2011}). As the blob regions are embedded within ocular dominance columns (\protect\hyperlink{ref-Horton1981}{Horton and Hubel, 1981}), they are also largely monocular (\protect\hyperlink{ref-Livingstone1984}{Livingstone and Hubel, 1984}; \protect\hyperlink{ref-Tychsen2004}{Tychsen et al., 2004}). More recent work has reported psychophysical evidence for unoriented chromatic (\protect\hyperlink{ref-Gheiratmand2013}{Gheiratmand et al., 2013}) and achromatic (\protect\hyperlink{ref-Meese2011}{Meese and Baker, 2011}) mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g.~Figures \ref{fig:BSratios} \& \ref{fig:TFdata}) owing to additivity of the electrophysiological response across independent monocular cells, rather than within binocular neurons. We therefore performed an additional analysis to investigate this possibility.

In the steady-state literature, one hallmark of a nonlinear system that pools inputs is the presence of intermodulation responses at the sums and differences of the fundamental flicker frequencies (\protect\hyperlink{ref-Tsai2012}{Tsai et al., 2012}). In Figure \ref{fig:IMfig} we plot the amplitude spectra of conditions from Experiment 1 in which the two eyes were stimulated at different frequencies (2Hz and 1.6Hz) but at the same contrast (48\%; these correspond to the binocular cross and dichoptic cross conditions in Figures \ref{fig:pupildata}d,e and \ref{fig:EEGdata}d,e). Consistent with the temporal properties of pupil responses and EEG activity, Figure \ref{fig:IMfig}a reveals a strong intermodulation difference response at 0.4Hz (red dashed line), and Figure \ref{fig:IMfig}b reveals an intermodulation sum response at 3.6Hz (red dashed line). The presence of these intermodulation terms is predicted by nonlinear gain control models of the type considered here (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Tsai2012}{Tsai et al., 2012}), and indicates that the processing of monocular flicker signals is not fully linear prior to the point at which they are combined across the eyes. Indeed, our model architecture makes specific predictions about the location of interocular suppression - it impacts before binocular combination, consistent with results from primate physiology (\protect\hyperlink{ref-Dougherty2019}{Dougherty et al., 2019}).

@@

\caption{Summary of intermodulation responses in pupillometry (a) and EEG (b) data. The data are pooled across the binocular cross and dichoptic cross conditions of Experiment 1, with a target contrast of 48\%. Vertical dashed lines indicate the fundamental flicker frequencies of 2Hz (F1; black) and 1.6Hz (F2; green), and the intermodulation difference (F1-F2 = 0.4Hz) and sum (F1+F2 = 3.6Hz) frequencies (red). Data are averaged across N = 30 participants, and shaded regions indicate ±1 standard error.}\label{fig:IMfig}
\end{figure}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Dougherty2019}{}}%
Dougherty K, Cox MA, Westerberg JA, Maier A. 2019. Binocular modulation of monocular V1 neurons. \emph{Curr Biol} \textbf{29}:381--391.e4. doi:\href{https://doi.org/10.1016/j.cub.2018.12.004}{10.1016/j.cub.2018.12.004}
